Ecosystem functional response across precipitation extremes in a sagebrush steppe

### 2017-08-25 manuscript/usses_water_ms.Rmd
final edits for coauthors

  **Methods.** We conducted a five year precipitation manipulation experiment in a sagebrush steppe ecosystem in Idaho, US. We used drought and irrigation treatments (approximately 50% decrease/increase) to investigate whether ecosystem functional response remains consistent at extreme precipitation levels. We recorded data on aboveground net primary productivity (ANPP), species abundance, and soil moisture. We fit a generalized linear mixed effects model to determine if the relationship between ANPP and soil moisture differed among treatments. We used nondimensional multivariate scaling to quantify community composition over the five years.

@@

We used non-dimensional multivariate scaling (NMDS) based on Bray-Curtis distances to see if community composition differed among treatments through time.
We first calculated Bray-Curtis distances among all plots for each year of the experiment and then extracted those distances for use in the NMDS.
Some values of standardized species' abundances were negative, which is not allowed for calculating Bray-Curtis distances.
We simply added '2' to each abundance value to ensure all values were greater than zero.
We plotted the first two axes of NMDS scores to see if community composition overlapped, or not, among treatments in each year.
We used the ```vegan::metaMDS()``` function [@Oksanen2016] to calculate Bray-Curtis distances and then to run the NMDS analysis.
We used the ```vegan::adonis()``` function [@Oksanen2016] to perform permutational multivariate analysis of variance to test whether treatment plots formed distinct groupings.
To test whether treatment plots were equally dispersed, or not, we used the ```vegan::betadisper()``` function [@Oksanen2016].

@@

Three of our five treatment years fell in years of below average rainfall (Fig. 2A).
Thus, those three years represent a lower magnitude of absolute change in precipitation experienced by the treatments.
ANPP varied from a minimum of 74.5 g m$^{-2}$ in 2014 to a maximum of 237.1 g m$^{-2}$ in 2016 when averaged across treatments (Fig. 2C).
ANPP was slightly higher in irrigation plots and slightly lower in drought plots (Fig. 2C), corresponding to estimated soil volumetric water content (VWC) differences among treatments (Fig. 2B).
Such differences in soil VWC indicate our treatment infrastructure was successful.

@@

Third, the life history traits of the dominant species in our study ecosystem may explain the consistently positive, but weak and uncertain, effect of soil moisture on ANPP (Fig. 3).
Species that live in variable environments, such as cold deserts, must have strategies to ensure long-term success as conditions vary.
One strategy is bet hedging, where species forego short-term gains to reduce the variance of long-term success [@Seger1987].
In other words, species do the same thing every year, with only minimal response to environmental conditions.
The dry and variable environment of the sagebrush steppe has likely selected for bet hedging species that can maintain function at low water availability and have weak responses to high water availability.
In so doing, the dominant species in our ecosystem avoid "boom and bust" cycles, which corresponds to the weak effect of soil moisture on ANPP (i.e., the Bayesian credible intervals for $\beta$ in Eq. 1 overlapping zero).

@@

The weaker soil moisture-ANPP relationship we observed for the irrigation treatment, evidenced by the less positive slope relative to controls (Fig. 3A, right panel), has three non-exclusive explanations.
First, species' biomass production may have hit physiological thresholds and/or become limited by other factors, such as nitrogen availability [@LeBauer2008].
\pba{PETER: DO WE KNOW ANYTHING ABOUT NITROGEN AT USSES? WANT TO ADD A SENTENCE HERE ABOUT HOW LIKELY OR UNLIKEL THIS IS...}
Second, the less positive slope could be another indicator of bet hedging in this community.
Avoiding high biomass production in wet years could benefit species in the long run if climate conditions are variable.
Third, the less positive slope may be a statistical artifact due the variability of ANPP within years and our low sample size.

>>>>

  **Methods.** We conducted a five year precipitation manipulation experiment in a sagebrush steppe ecosystem in Idaho, United States. We used drought and irrigation treatments (approximately 50% decrease/increase) to investigate whether ecosystem functional response remains consistent at extreme precipitation levels. We recorded data on aboveground net primary productivity (ANPP), species abundance, and soil moisture. We fit a generalized linear mixed effects model to determine if the relationship between ANPP and soil moisture differed among treatments. We used nonmetric multidimensional scaling to quantify community composition over the five years.

@@

We used nonmetric multidimensional scaling (NMDS) based on Bray-Curtis distances to see if community composition differed among treatments through time.
We first calculated Bray-Curtis distances among all plots for each year of the experiment and then extracted those distances for use in the NMDS.
Some values of standardized species' abundances were negative, which is not allowed for calculating Bray-Curtis distances.
We simply added '2' to each abundance value to ensure all values were greater than zero.
We plotted the first two axes of NMDS scores to see if community composition overlapped, or not, among treatments in each year.
We used the ```vegan::metaMDS()``` function [@Oksanen2016] to calculate Bray-Curtis distances and then to run the NMDS analysis.
We used the ```vegan::adonis()``` function [@Oksanen2016] to perform permutational multivariate analysis of variance to test whether treatment plots formed distinct groupings.
To test whether treatment plots were equally dispersed, or not, we used the ```vegan::betadisper()``` function [@Oksanen2016].

@@

Three of our five treatment years fell in years of below average rainfall (Fig. 2A).
Thus, those three years represent a lower magnitude of absolute change in precipitation experienced by the treatments.
ANPP varied from a minimum of 74.5 g m$^{-2}$ in 2014 to a maximum of 237.1 g m$^{-2}$ in 2016 when averaged across treatments (Fig. 2C).
ANPP was slightly higher in irrigation plots and slightly lower in drought plots (Fig. 2C), corresponding to estimated soil volumetric water content (VWC) differences among treatments (Fig. 2B).
Such differences in soil VWC indicate our treatment infrastructure was successful.
ANPP was highly variable within years (Fig. 2C).

@@

Third, the life history traits of the dominant species in our study ecosystem may explain the consistently positive, but weak and uncertain, effect of soil moisture on ANPP (Fig. 3).
Species that live in variable environments, such as cold deserts, must have strategies to ensure long-term success as conditions vary.
One strategy is bet hedging, where species forego short-term gains to reduce the variance of long-term success [@Seger1987].
In other words, species do the same thing every year, with only minimal response to environmental conditions.
The dry and variable environment of the sagebrush steppe has likely selected for bet hedging species that can maintain function at low water availability and have weak responses to high water availability.
In so doing, the dominant species in our ecosystem avoid "boom and bust" cycles, which corresponds to the weak effect of soil moisture on ANPP (i.e., the Bayesian credible intervals for the slopes in $\boldsymbol{\beta}$ in Eq. 1 overlapping zero).

@@

The weaker soil moisture-ANPP relationship we observed for the irrigation treatment, evidenced by the less positive slope relative to controls (Fig. 3A, right panel), has three non-exclusive explanations.
First, species' biomass production may have hit physiological thresholds and/or become limited by other factors, such as nitrogen availability [@LeBauer2008].
\pba{PETER: DO WE KNOW ANYTHING ABOUT NITROGEN AT USSES? WANT TO ADD A SENTENCE HERE ABOUT HOW LIKELY OR UNLIKEL THIS IS...}
Second, the less positive slope could be another indicator of bet hedging in this community.
Avoiding high biomass and seed production in wet years could benefit species in the long run if climate conditions are variable.
Third, the less positive slope may be a statistical artifact due the variability of ANPP within years and our low sample size.

### 2017-08-25 manuscript/usses_water_ms.tex
final edits for coauthors

\textbf{Methods.} We conducted a five year precipitation manipulation
experiment in a sagebrush steppe ecosystem in Idaho, US. We used drought
and irrigation treatments (approximately 50\% decrease/increase) to
investigate whether ecosystem functional response remains consistent at
extreme precipitation levels. We recorded data on aboveground net
primary productivity (ANPP), species abundance, and soil moisture. We
fit a generalized linear mixed effects model to determine if the
relationship between ANPP and soil moisture differed among treatments.
We used nondimensional multivariate scaling to quantify community
composition over the five years.

@@

We used non-dimensional multivariate scaling (NMDS) based on Bray-Curtis
distances to see if community composition differed among treatments
through time. We first calculated Bray-Curtis distances among all plots
for each year of the experiment and then extracted those distances for
use in the NMDS. Some values of standardized species' abundances were
negative, which is not allowed for calculating Bray-Curtis distances. We
simply added `2' to each abundance value to ensure all values were
greater than zero. We plotted the first two axes of NMDS scores to see
if community composition overlapped, or not, among treatments in each
year. We used the \texttt{vegan::metaMDS()} function \citep{Oksanen2016}
to calculate Bray-Curtis distances and then to run the NMDS analysis. We
used the \texttt{vegan::adonis()} function \citep{Oksanen2016} to
perform permutational multivariate analysis of variance to test whether
treatment plots formed distinct groupings. To test whether treatment
plots were equally dispersed, or not, we used the
\texttt{vegan::betadisper()} function \citep{Oksanen2016}.

@@

Three of our five treatment years fell in years of below average
rainfall (Fig. 2A). Thus, those three years represent a lower magnitude
of absolute change in precipitation experienced by the treatments. ANPP
varied from a minimum of 74.5 g m\(^{-2}\) in 2014 to a maximum of 237.1
g m\(^{-2}\) in 2016 when averaged across treatments (Fig. 2C). ANPP was
slightly higher in irrigation plots and slightly lower in drought plots
(Fig. 2C), corresponding to estimated soil volumetric water content
(VWC) differences among treatments (Fig. 2B). Such differences in soil
VWC indicate our treatment infrastructure was successful.

@@

Third, the life history traits of the dominant species in our study
ecosystem may explain the consistently positive, but weak and uncertain,
effect of soil moisture on ANPP (Fig. 3). Species that live in variable
environments, such as cold deserts, must have strategies to ensure
long-term success as conditions vary. One strategy is bet hedging, where
species forego short-term gains to reduce the variance of long-term
success \citep{Seger1987}. In other words, species do the same thing
every year, with only minimal response to environmental conditions. The
dry and variable environment of the sagebrush steppe has likely selected
for bet hedging species that can maintain function at low water
availability and have weak responses to high water availability. In so
doing, the dominant species in our ecosystem avoid ``boom and bust''
cycles, which corresponds to the weak effect of soil moisture on ANPP
(i.e., the Bayesian credible intervals for \(\beta\) in Eq. 1
overlapping zero).

@@

The weaker soil moisture-ANPP relationship we observed for the
irrigation treatment, evidenced by the less positive slope relative to
controls (Fig. 3A, right panel), has three non-exclusive explanations.
First, species' biomass production may have hit physiological thresholds
and/or become limited by other factors, such as nitrogen availability
\citep{LeBauer2008}.
\pba{PETER: DO WE KNOW ANYTHING ABOUT NITROGEN AT USSES? WANT TO ADD A SENTENCE HERE ABOUT HOW LIKELY OR UNLIKEL THIS IS...}
Second, the less positive slope could be another indicator of bet
hedging in this community. Avoiding high biomass production in wet years
could benefit species in the long run if climate conditions are
variable. Third, the less positive slope may be a statistical artifact
due the variability of ANPP within years and our low sample size.

>>>>

\textbf{Methods.} We conducted a five year precipitation manipulation
experiment in a sagebrush steppe ecosystem in Idaho, United States. We
used drought and irrigation treatments (approximately 50\%
decrease/increase) to investigate whether ecosystem functional response
remains consistent at extreme precipitation levels. We recorded data on
aboveground net primary productivity (ANPP), species abundance, and soil
moisture. We fit a generalized linear mixed effects model to determine
if the relationship between ANPP and soil moisture differed among
treatments. We used nonmetric multidimensional scaling to quantify
community composition over the five years.

@@

We used nonmetric multidimensional scaling (NMDS) based on Bray-Curtis
distances to see if community composition differed among treatments
through time. We first calculated Bray-Curtis distances among all plots
for each year of the experiment and then extracted those distances for
use in the NMDS. Some values of standardized species' abundances were
negative, which is not allowed for calculating Bray-Curtis distances. We
simply added `2' to each abundance value to ensure all values were
greater than zero. We plotted the first two axes of NMDS scores to see
if community composition overlapped, or not, among treatments in each
year. We used the \texttt{vegan::metaMDS()} function \citep{Oksanen2016}
to calculate Bray-Curtis distances and then to run the NMDS analysis. We
used the \texttt{vegan::adonis()} function \citep{Oksanen2016} to
perform permutational multivariate analysis of variance to test whether
treatment plots formed distinct groupings. To test whether treatment
plots were equally dispersed, or not, we used the
\texttt{vegan::betadisper()} function \citep{Oksanen2016}.

@@

Three of our five treatment years fell in years of below average
rainfall (Fig. 2A). Thus, those three years represent a lower magnitude
of absolute change in precipitation experienced by the treatments. ANPP
varied from a minimum of 74.5 g m\(^{-2}\) in 2014 to a maximum of 237.1
g m\(^{-2}\) in 2016 when averaged across treatments (Fig. 2C). ANPP was
slightly higher in irrigation plots and slightly lower in drought plots
(Fig. 2C), corresponding to estimated soil volumetric water content
(VWC) differences among treatments (Fig. 2B). Such differences in soil
VWC indicate our treatment infrastructure was successful. ANPP was
highly variable within years (Fig. 2C).

@@

Third, the life history traits of the dominant species in our study
ecosystem may explain the consistently positive, but weak and uncertain,
effect of soil moisture on ANPP (Fig. 3). Species that live in variable
environments, such as cold deserts, must have strategies to ensure
long-term success as conditions vary. One strategy is bet hedging, where
species forego short-term gains to reduce the variance of long-term
success \citep{Seger1987}. In other words, species do the same thing
every year, with only minimal response to environmental conditions. The
dry and variable environment of the sagebrush steppe has likely selected
for bet hedging species that can maintain function at low water
availability and have weak responses to high water availability. In so
doing, the dominant species in our ecosystem avoid ``boom and bust''
cycles, which corresponds to the weak effect of soil moisture on ANPP
(i.e., the Bayesian credible intervals for the slopes in
\(\boldsymbol{\beta}\) in Eq. 1 overlapping zero).

@@

The weaker soil moisture-ANPP relationship we observed for the
irrigation treatment, evidenced by the less positive slope relative to
controls (Fig. 3A, right panel), has three non-exclusive explanations.
First, species' biomass production may have hit physiological thresholds
and/or become limited by other factors, such as nitrogen availability
\citep{LeBauer2008}.
\pba{PETER: DO WE KNOW ANYTHING ABOUT NITROGEN AT USSES? WANT TO ADD A SENTENCE HERE ABOUT HOW LIKELY OR UNLIKEL THIS IS...}
Second, the less positive slope could be another indicator of bet
hedging in this community. Avoiding high biomass and seed production in
wet years could benefit species in the long run if climate conditions
are variable. Third, the less positive slope may be a statistical
artifact due the variability of ANPP within years and our low sample
size.
